Thank you to my parents, for teaching me, for encouraging me to pursue my own education, and putting up with the huge piles of messed-about-with hardware that follow me around. Your unconditional love and support has been a comfort to me in times of stress, and a joy always.

Thank you to Dr. Jay McCarthy for the useful pointers in programming languages and verification, and for keeping me on track when I was getting into the weeds of Turing completeness. 

Thank you to Dr. Radhika Nagpal for the loan of the E-Pucks, for taking the time to be on my committee, and for an impressive body of work in swarm robotics that inspired and informed some of this work. 

Thank you to Dr. Holly Yanco, for encouraging me to turn one of my side projects into a dissertation, for providing guidance and resources for all these years, and finally for the encouragement to get out. Your ceaseless work to run the Robotics Lab and the NERVE Center and encourage a good mixture of freedom and focus for the people working there have made it a fantastic place to work. 

Thank you to Jonathan Roche for vastly speeding up the virtual laser, to James Kuczynski and Dalton Curtin for their attention to detail in coding the user responses, and everyone else in the UML robotics lab that I've asked to read papers, write ROS modules, or otherwise help out. Collaboration is one of our lab's big strengths, keep it that way when I'm gone.  

%Thank you to Vanessa Evers, Edwin Dertien, and the staff of the Human Media Interaction group at the University of Twente for hosting me while I worked on the swarm hardware. It was engaging and inspiring to ``do things with stuff'' with you. 
