\chapter{Timeline}

As of this writing, the deadlines for filing Declaration of Intent to Graduate forms for spring completion is unknown, but is likely in the last week of March or first week of April. 
The deadline for submitting a thesis or dissertation is May 3rd of 2018. 
Commencement is May 19th. 

\section{December 2017}

\subsection{Week 1}

Code first 10 users with two coders. A coding program was written to facilitate coding, based on the codes present in the first 10 users. Use of the coding program will ease automation of parts of the data analysis. 
Check Cohen's K on subset of data.
Write codebook for full coding. 

\subsection{Week 2}

Recode full data set using codebook.
Start writing data analysis scripts.

\subsection{Week 3}

Christmas travel. 
Data analysis.

\subsection{Week 4}

Christmas. 
Analysis of results from coding.
Design validation experiment and submit to IRB.

\section{January 2018}

\subsection{Week 1}

Write up results of data analysis.
At this point, it should be clear that there either is or is not a transition point where the user changes the gestures they use for a bigger swarm. 
If there is no transition, then it would seem to indicate that the same UI can be used for individual robots and swarms. 

\subsection{Week 2}

Modify state machine to recognize most common gestures, as determined from coding and analysis. 
Start implementing validation experiment in simulation. 

\subsection{Week 3}

Implement or modify state machine to emit proper GCPR for common commands. 
Connect GCPR and GCPR interpreter to real robots.
 
\subsection{Week 4}

Test full gesture-to-robot compilation path. Components of this chain have already been at least partly developed, including a basic gesture recognizer, a state machine for detecting commands based on gestures, and a GCPR evaluator for ROS.

\section{February 2018}

\subsection{Week 1}

Debug complete system. 

\subsection{Week 2}

Debug system. 
Shoot video of system when operational, or record rosbags. 

\subsection{Week 3}

Debug compiler and robot runner. 
Write up full interface/system, including instructions for other people to operate it. 
If enough of system is operational, do validation experiment on real system, connect to simulation and do it in simulation. 

\subsection{Week 4}

Merge experiment write-up paper and system documentation with thesis.
Start presentation for defence, based on presentation for proposal. 
Either work on simulation for validation or running of validation experiment on physical system. 

\section{March 2018}

Defence should be scheduled for this month or April. 

\subsection{Week 1}

Writing.
Finish validation experiment.
Begin analysis of validation experiment. 

\subsection{Week 2}

Writing.
Analysis of validation experiment. 
Begin write-up of validation experiment. 
Submit intention to graduate form. 

\subsection{Week 3}

Writing, finish write-up of validation experiment. Merge validation experiment write-up with thesis. 

\subsection{Week 4}

Writing. 
Deadline to submit intention to graduate form is likely this week. 

\section{April 2018}

If defence hasn't been done already, do it this month. 

\subsection{Week 1}

Finish writing, by this point there should be papers on the gesture discovery experiment, the swarm itself, and the validation experiment. 

\subsection{Week 2}

Finish writing, polish. 

\subsection{Week 3}

Ideally submit to library this week.

\subsection{Week 4}

Deadline to submit thesis to library is May 3rd.